\def\duedate{\today}
\def\HWnum{10}
\input{../preamble.tex}

\begin{document}

\prob{1}{

Consider the Pauli matrix:
\begin{eqnarray}
    \sigma_{y} = \begin{pmatrix}
        0 & -i \\
        i & 0
    \end{pmatrix}
.\end{eqnarray}

(a) Obtain the eigenvalues and eigenvectors of $\sigma_{y}$.

(b) Find a unitary operator $U$ which diagonalizes $\sigma_{y}$.

}

\sol{

(a) We can easily see that the characteristic equation of $\sigma_{y}$ is
\begin{eqnarray}
    \lambda^2 - 1 = 0 \Rightarrow \lambda = \pm 1
.\end{eqnarray}
Thus, the components of the eigenvectors satisfy
\begin{eqnarray}
    -\lambda a_1 = i a_2 \Rightarrow a_2 = i \lambda a_1
,\end{eqnarray}
and the eigenvectors are then (up to some constant)
\begin{eqnarray}
\eqbox{
\begin{aligned}
    \begin{pmatrix}
        1 \\ i
    \end{pmatrix}
    &\leftrightarrow 1
    \\
    \begin{pmatrix}
        1 \\ -i
    \end{pmatrix}
    &\leftrightarrow -1
\end{aligned}
}
.\end{eqnarray}

(b) We can easily diagonalize $\sigma_{y}$ by writing
\begin{eqnarray}
\eqbox{
    U = \frac{1}{\sqrt{2}} \begin{pmatrix}
        1 & 1 \\
        i & -i
    \end{pmatrix}
}
,\end{eqnarray}
where
\begin{eqnarray}
    A' = U^{\dagger} A U = \begin{pmatrix}
        1 & 0 \\
        0 & -1
    \end{pmatrix}
.\end{eqnarray}

}


\prob{2}{

Consider a rotation matrix $a$ with complex elements $a_{ij}$:
\begin{eqnarray}
    a = \begin{pmatrix}
        a_{11} & a_{12} & a_{13} \\
        a_{21} & a_{22} & a_{23} \\
        a_{31} & a_{32} & a_{33}
    \end{pmatrix}
.\end{eqnarray}
Find the inverse matrix $a^{-1}$.

}

\sol{

Since $a$ is a rotation matrix, it must be unitary.
We prove this as follows.
Let us posit a defining property of rotation matrices: they preserve the norm of a vector under a transformation.
That is, $||v|| = ||R v ||$.
Using this we see that
\begin{eqnarray}
    v^{\dagger} v = v^{\dagger} ( R^{\dagger} R ) v \Rightarrow R^{\dagger} R = \mathbb{I}
.\end{eqnarray}

Another way to see this is by recalling from Quantum Mechanics that the angular momentum operator $J$ is the generator of rotations.
That is, we can define a generic rotation operator through an angle $\theta$ about an axis defined by a unit vector $\vu*{n}$ as $R = \exp(-i \theta \vu*{n} \cdot \va*{J})$, which has inverse $R^{-1} = R^{\dagger}$.
This is a more abstract object than a matrix in general: it performs a rotation of states $\ket{\psi} \rightarrow \ket{\psi'} = R \ket{\psi}$ in the Hilbert space.
We can, however, typically represent the operator $R$ using a matrix, and if our Hilbert space has dimension 3 (e.g. a spin-1 system), then our rotation matrix $R \in \mathbb{C}^{3 \times 3}$.

Finally, we write down explicitly
\begin{eqnarray}
\eqbox{
    a^{-1} = a^{\dagger} = \begin{pmatrix}
        a_{11}^{*} & a_{21}^{*} & a_{31}^{*} \\
        a_{12}^{*} & a_{22}^{*} & a_{32}^{*} \\
        a_{13}^{*} & a_{23}^{*} & a_{33}^{*}
    \end{pmatrix}
}
.\end{eqnarray}

}


\prob{3}{

Using the definition of a function of a matrix given in Lect. 22 and the properties of the Pauli matrices $\sigma_{i}^2 = \mathbb{I}$, prove the following relation:
\begin{eqnarray}
    \exp(i \theta \sigma_{z}) = \mathbb{I} \cos{\theta} + i \sigma_{z} \sin{\theta}
.\end{eqnarray}

}

\sol{

From the definition of the exponential of an operator, we can write
\begin{eqnarray}
    \exp(i \theta \sigma_{z}) = \sum_{n=0}^{\infty} \frac{1}{n!} (i \theta \sigma_{z})^{n}
.\end{eqnarray}
Let us split the sum over even and odd $n$ such that
\begin{eqnarray}
\eqbox{
\begin{aligned}
    \exp(i \theta \sigma_{z}) &= \sum_{n=0}^{\infty} \frac{1}{(2n)!} (i \theta \sigma_{z})^{2n} + \sum_{n=0}^{\infty} \frac{1}{(2n+1)!} (i \theta \sigma_{z})^{2n+1} \\
                              &= \sum_{n=0}^{\infty} \frac{(-1)^{n}}{(2n)!} \theta^{2n} \mathbb{I} + i \sum_{n=0}^{\infty} \frac{(-1)^{n}}{(2n+1)!} \theta^{2n+1} \sigma_{z} \\
                              &= \cos{\theta} \mathbb{I} + i \sin{\theta} \sigma_{z}
\end{aligned}
}
.\end{eqnarray}


}



\end{document}
