\chapter{Spherical Harmonics}

This section will be fairly quick since in the previous chapter, we have already laid down most of the foundation.
The spherical harmonics are functions $Y(\theta,\phi)$ which satisfy Laplace's equation, where $\theta$ and $\phi$ are the azimuthal and polar angles, respectively, in spherical coordinates:
\begin{eqnarray}
    r^2 \laplacian Y(\theta,\phi) = \frac{1}{\sin{\theta}} \pdv{\theta} \Big( \sin{\theta} \pdv{Y(\theta,\phi)}{\theta} \Big) + \frac{1}{\sin^2{\theta}} \pdv[2]{Y(\theta,\phi)}{\phi} = 0
.\end{eqnarray}
In the last chapter, we proceeded to solve this equation via a separable ansatz $Y(\theta,\phi) = P(\theta) T(\phi)$ and found that the azimuthal dependence is given by the associated Legendre polynomials $P(\theta) = P_{l}^{m}(\cos{\theta})$ and the polar dependence is given by phase factors $e^{i m \phi}$, where $l \geq 0$ and $-l \leq m \leq l$ are integers.
The full separable solution is then just
\begin{eqnarray}
    Y_{lm}(\theta,\phi) = \mathcal{N}_{lm} P_{l}^{m}(\cos{\theta}) e^{im\phi}
.\end{eqnarray}
The normalization factor is fixed by requiring (as is conventional) that
\begin{eqnarray}
    \int \dd{\Omega} Y_{l'm'}^{*}(\theta,\phi) Y_{lm}^{*}(\theta,\phi) = \delta_{l'l} \delta_{m'm}
.\end{eqnarray}
The orthogonality is already a given with the associated Legendre polynomials and phase factors.
We can determine the normalization factor quite easily then:
\begin{align}
    1 = \int_{0}^{2\pi} \dd{\phi} \int_{0}^{\pi} \dd{\theta} \sin{\theta} |Y_{lm}(\theta,\phi)|^2 &= \mathcal{N}_{lm}^2 (2 \pi) \int_{0}^{\pi} \dd{\theta} \sin{\theta} [P_{l}^{m}(\cos{\theta})]^2 \nonumber \\
                                                                                              &= \mathcal{N}_{lm}^2 (2 \pi) \int_{-1}^{1} \dd{x} [ P_{l}^{m}(x) ]^2 \nonumber \\
                                                                                              &= \mathcal{N}_{lm} \frac{4 \pi}{2l+1} \frac{(l+m)!}{(l-m)!}
.\end{align}
Solving for $\mathcal{N}_{lm}$ then gives the normalized spherical harmonics as
\begin{eqnarray}
    Y_{lm}(\theta,\phi) = \sqrt{\frac{2l+1}{4 \pi} \frac{(l-m)!}{(l+1)!}} P_{l}^{m}(\cos{\theta}) e^{im\phi}
,\end{eqnarray}
and that is the spherical harmonics in a nutshell.

One final note is about the completeness of the spherical harmonics.
Consistent with what we claimed in the Sturm-Liouville section, for any square-integrable function $f(\theta,\phi)$, we can expand it in spherical harmonics as
\begin{eqnarray}
    f(\theta,\phi) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} f_{lm} Y_{lm}(\theta,\phi)
,\end{eqnarray}
where the coefficients are given through the orthonormality condition
\begin{eqnarray}
    f_{lm} = \int \dd{\Omega} Y_{lm}^{*}(\theta,\phi) f(\theta,\phi)
.\end{eqnarray}

If our function is rotationally symmetric about the $z$-axis then $f(\theta,\phi) = f(\theta)$ and only the $m = 0$ term survives, yielding an expansion in Legendre polynomials:
\begin{eqnarray}
    f(\theta) = \sum_{l=0}^{\infty} f_{l} P_{l}(\cos{\theta})
\end{eqnarray}
with some constants absorbed into the coefficients
\begin{eqnarray}
    f_{l} = \frac{2l+1}{2} \int_{0}^{\pi} \dd{\theta} \sin{\theta} P_{l}(\cos{\theta}) f(\theta)
.\end{eqnarray}






